---
title: "GSS challange"
author: "Taylor Stohl"
date: "Summer 2021"
output: html_document
---

```{r include=FALSE,echo=FALSE}
require(tidyverse)
require(tigerstats)
require(knitr)
GSS <-read.csv(file="GSScsv.csv",header=TRUE)
library(dplyr)
GSS <- GSS %>%
mutate(across(where(is.character), as.factor))
```

# Introduction

<!--In this section you explain what you are trying to show.  Where did the data come from?  What is the research or other question you are trying to answer?.  If you skimp in this section expect to be marked down in your grade. Paint a picture of the project a couple of paragraphs is expected!-->

Is attitude toward capitol punishment (CAPPUN) independent of race (RACE)?
If not, then whose view differs from whom??

# Methods

<!--Decide on your methods:  use "variable analysis" or other appropriate descriptors.  Make sure to choose at least one graphical method and at least one numerical method. Discuss what data manipulations may be needed for example subsetting or other binning procedures and what sort of graphical displays and inferential tools will be used!-->

# Results

<!--Divide this section into two sub-sections:  One for your descriptive  results and one for your inferential results.!-->

## Descriptive Results

### Graphical Descriptive Results

<!--Graphical results here.  Make sure to show your code.  Provide appropriate labels for axes, giving units if possible, and provide a good title for the graph, too.  Use the graphical results to describe the patterns if any that exist in the data as focused toward the research question. Bottom line is to explain the features of the graph that are relevant to conforming or departing to what is expected by your null hypothesis !-->


```{r}
barchartGC(~RACE + CAPPUN,data=GSS)
barchartGC(~RACE + CAPPUN,data=GSS, type="percent")

```


In the first plot, the number of participants in each cetegory are displayed. There are about 30 participants that fall into the Black category, around 24 people in the Other category and lastly, around 120 people in the White category. Most of the participants in this study fell into the white category.

Based on what can be seen in the second plot, it appears that the black population is far more against CAPPUN(capitol punishment), than the white population and the Other category. While around 38% of the black population was in favor of CAPPUN, around the same percentage of the white population( around 38%) was not opposed CAPPUN. The Bar representing "Other," refers to races that are not black or white. These results seem to be vastly different  for the results of the black community as well. Of the other category,it looks like about 80% are in favor of CAPPUN and only about 20% are against it. All in all, The black community is much more against CAPPUN(capitol punishment) than the Other community and the white population. The white group and other group have results that seem to be more closely related.


### Numerical Descriptive Results

<!--Numerical results go here. Use the numerical results to describe the patterns if any that exist in the data as focused toward the research question!-->

```{r}
table1 <- xtabs(~RACE + CAPPUN, data=GSS)
rowPerc(table1)
colPerc(table1)
```

## Inferential Results

<!--State hypothesis clearly.  Make sure your discussion of the inferential test covers all the aspects that the test output produces, such as test statistic, p-value etc.  Make a decision about the null hypothesis, explain the assumptions on which the selected test/procedure was based, and why the chosen procedure satisfys the assumptions and is appropriate to answer the research question!-->

```{r}
chisq.test(table1)
print("killroy was here")
chisqtestGC(table1)
```









# Discussion and Conclusion

<!--Discussion and conclusion here.  If you found a relationship be sure to consider whether the relationship occurs because one of the variavbles causes the other, or whether they perhasps are related for some other reason.  Watch the chapter 6 videos from the GeorgeTown videos collection.!-->
